---
title: "Clinical Characteristics"
output: html_document
date: "2024-10-22"
---

# Prepare environment & Load Data
```{r}
# Clean environment
rm( list = ls() )

# load libraries
library(xlsx)
library(tidyverse)
library(khroma)
library(colorBlindness)
library(ggplot2)

# load data
setwd("~/GitHub/BCH-GBS/2.Statistical_Analysis/Data")
data <- read.xlsx("Clinical_Genomic_data.xlsx", sheetIndex = 1) 
# Remove contaminated samples
data_clean <- data|>
  filter(! ID %in% c("1_S1","32_S32","35_S35","81_S81"))|> 
  mutate(ICU = ifelse(Admission_loc=="ICU", 1,0))

# Subset to infants alone 
data_clean_infants <- data_clean |> filter(days_at_dx <= 365)
```

# 1. Clinical Correlates of Severe Disease 
## 1.0. Function to calculate Odds Ratio and 95% CI
```{r}
calculate_or_ci <- function(model) {
  coef_logit <- coef(model)[2]
  se_logit <- coef(summary(model))[2, "Std. Error"]
  lower_ci <- coef_logit - 1.96 * se_logit
  upper_ci <- coef_logit + 1.96 * se_logit
  
  odds_ratio <- exp(coef_logit)
  lower_or <- exp(lower_ci)
  upper_or <- exp(upper_ci)
  
  or_ci <- paste0(round(odds_ratio, 2), " (", round(lower_or, 2), "-", round(upper_or, 2), ")")
  return(or_ci)
}
```
## 1.1. Infant disease
```{r}
# Logistic regression models
model_infant_Hb            <- glm(infant ~ Hb,data_clean, family = binomial)
model_infant_WBC           <- glm(infant ~ WBC,data_clean, family = binomial)
model_infant_Leukocytosis  <- glm(infant ~ Leukocytosis,data_clean, family = binomial)
model_infant_Leukopenia    <- glm(infant ~ Leukopenia,data_clean, family = binomial)
model_infant_platelet      <- glm(infant ~ platelet,data_clean, family = binomial)
model_infant_ANC           <- glm(infant ~ ANC,data_clean, family = binomial)
model_infant_Neutropenia   <- glm(infant ~ Neutropenia,data_clean, family = binomial)
# Calculate OR and 95% CI for each factor
or_infant_Hb           <- calculate_or_ci(model_infant_Hb)
or_infant_WBC          <- calculate_or_ci(model_infant_WBC)
or_infant_Leukocytosis <- calculate_or_ci(model_infant_Leukocytosis)
or_infant_Leukopenia   <- calculate_or_ci(model_infant_Leukopenia)
or_infant_platelet     <- calculate_or_ci(model_infant_platelet)
or_infant_ANC          <- calculate_or_ci(model_infant_ANC)
or_infant_Neutropenia  <- calculate_or_ci(model_infant_Neutropenia)
# Create a p-value vector from the model_infants
p_val_infant_Hb            <- round(coef(summary(model_infant_Hb))[2,4], 3)
p_val_infant_WBC           <- round(coef(summary(model_infant_WBC))[2,4], 3)
p_val_infant_Leukocytosis  <- round(coef(summary(model_infant_Leukocytosis))[2,4], 3)
p_val_infant_Leukopenia    <- round(coef(summary(model_infant_Leukopenia))[2,4], 3)
p_val_infant_platelet      <- round(coef(summary(model_infant_platelet))[2,4], 3)
p_val_infant_ANC           <- round(coef(summary(model_infant_ANC))[2,4], 3)
p_val_infant_Neutropenia   <- round(coef(summary(model_infant_Neutropenia))[2,4], 3)
# Create the summary table
Table1_1 <- data_clean |>
  mutate(infant = factor(infant, levels = c(0, 1))) |>
  group_by(infant) |>
  summarise(N = n(),
            Hb_mean_SD = paste(round(mean(Hb, na.rm = TRUE), 2), " (", round(sd(Hb, na.rm = TRUE), 2), ")", sep = ""),
            WBC_mean_SD = paste(round(mean(WBC, na.rm = TRUE), 2), " (", round(sd(WBC, na.rm = TRUE), 2), ")", sep = ""),
            Leukocytosis_Yes_p = paste(sum(Leukocytosis == 1, na.rm = TRUE), " (", round(100 * mean(Leukocytosis == 1, na.rm = TRUE), 1), ")", sep = ""),
            Leukopenia_Yes_p = paste(sum(Leukopenia == 1, na.rm = TRUE), " (", round(100 * mean(Leukopenia == 1, na.rm = TRUE), 1), ")", sep = ""),
            Platelet_mean_SD = paste(round(mean(platelet, na.rm = TRUE), 2), " (", round(sd(platelet, na.rm = TRUE), 2), ")", sep = ""),
            ANC_mean_SD = paste(round(mean(ANC, na.rm = TRUE), 2), " (", round(sd(ANC, na.rm = TRUE), 2), ")", sep = ""),
            Neutropenia_Yes_p = paste(sum(Neutropenia == 1, na.rm = TRUE), " (", round(100 * mean(Neutropenia == 1, na.rm = TRUE), 1), ")", sep = "")) |>
  mutate(Value = paste0(infant, "(N=", N, ")"), .before = Hb_mean_SD) |>
  select(-infant, -N)
# Add OR row directly during table creation
or_infant_row <- c("OR (95% CI)", or_infant_Hb, or_infant_WBC, or_infant_Leukocytosis, or_infant_Leukopenia, or_infant_platelet, or_infant_ANC, or_infant_Neutropenia)
# Add p-value row
p_val_infant_row <- c("p-value", p_val_infant_Hb, p_val_infant_WBC, p_val_infant_Leukocytosis, p_val_infant_Leukocytosis, p_val_infant_platelet, p_val_infant_ANC, p_val_infant_Neutropenia)
# Bind both rows to the table
Table1_1 <- rbind(Table1_1, or_infant_row, p_val_infant_row)
# View the final table
Table1_1
```
## 1.2. Age of onset among infants: LOD vs VLOD
```{r}
data_clean_LOD <- data_clean_infants |>
  filter(Age_cat !="EOD")|>
  mutate(LOD= ifelse(Age_cat =="LOD", 1, 0))
# Logistic regression models
model_LOD_Hb            <- glm(LOD ~ Hb,data_clean_LOD, family = binomial)
model_LOD_WBC           <- glm(LOD ~ WBC,data_clean_LOD, family = binomial)
model_LOD_Leukocytosis  <- glm(LOD ~ Leukocytosis,data_clean_LOD, family = binomial)
model_LOD_Leukopenia    <- glm(LOD ~ Leukopenia,data_clean_LOD, family = binomial)
model_LOD_platelet      <- glm(LOD ~ platelet,data_clean_LOD, family = binomial)
model_LOD_ANC           <- glm(LOD ~ ANC,data_clean_LOD, family = binomial)
model_LOD_Neutropenia   <- glm(LOD ~ Neutropenia,data_clean_LOD, family = binomial)
# Calculate OR and 95% CI for each factor
or_LOD_Hb           <- calculate_or_ci(model_LOD_Hb)
or_LOD_WBC          <- calculate_or_ci(model_LOD_WBC)
or_LOD_Leukocytosis <- calculate_or_ci(model_LOD_Leukocytosis)
or_LOD_Leukopenia   <- calculate_or_ci(model_LOD_Leukopenia)
or_LOD_platelet     <- calculate_or_ci(model_LOD_platelet)
or_LOD_ANC          <- calculate_or_ci(model_LOD_ANC)
or_LOD_Neutropenia  <- calculate_or_ci(model_LOD_Neutropenia)
# Create a p-value vector from the model_LODs
p_val_LOD_Hb            <- round(coef(summary(model_LOD_Hb))[2,4], 3)
p_val_LOD_WBC           <- round(coef(summary(model_LOD_WBC))[2,4], 3)
p_val_LOD_Leukocytosis  <- round(coef(summary(model_LOD_Leukocytosis))[2,4], 3)
p_val_LOD_Leukopenia    <- round(coef(summary(model_LOD_Leukopenia))[2,4], 3)
p_val_LOD_platelet      <- round(coef(summary(model_LOD_platelet))[2,4], 3)
p_val_LOD_ANC           <- round(coef(summary(model_LOD_ANC))[2,4], 3)
p_val_LOD_Neutropenia   <- round(coef(summary(model_LOD_Neutropenia))[2,4], 3)
# Create the summary table
Table1_2 <- data_clean_LOD |>
  mutate(LOD = factor(LOD, levels = c(0, 1))) |>
  group_by(LOD) |>
  summarise(N = n(),
            Hb_mean_SD = paste(round(mean(Hb, na.rm = TRUE), 2), " (", round(sd(Hb, na.rm = TRUE), 2), ")", sep = ""),
            WBC_mean_SD = paste(round(mean(WBC, na.rm = TRUE), 2), " (", round(sd(WBC, na.rm = TRUE), 2), ")", sep = ""),
            Leukocytosis_Yes_p = paste(sum(Leukocytosis == 1, na.rm = TRUE), " (", round(100 * mean(Leukocytosis == 1, na.rm = TRUE), 1), ")", sep = ""),
            Leukopenia_Yes_p = paste(sum(Leukopenia == 1, na.rm = TRUE), " (", round(100 * mean(Leukopenia == 1, na.rm = TRUE), 1), ")", sep = ""),
            Platelet_mean_SD = paste(round(mean(platelet, na.rm = TRUE), 2), " (", round(sd(platelet, na.rm = TRUE), 2), ")", sep = ""),
            ANC_mean_SD = paste(round(mean(ANC, na.rm = TRUE), 2), " (", round(sd(ANC, na.rm = TRUE), 2), ")", sep = ""),
            Neutropenia_Yes_p = paste(sum(Neutropenia == 1, na.rm = TRUE), " (", round(100 * mean(Neutropenia == 1, na.rm = TRUE), 1), ")", sep = "")) |>
  mutate(Value = paste0(LOD, "(N=", N, ")"), .before = Hb_mean_SD) |>
  select(-LOD, -N)
# Add OR row directly during table creation
or_LOD_row <- c("OR (95% CI)", or_LOD_Hb, or_LOD_WBC, or_LOD_Leukocytosis, or_LOD_Leukopenia, or_LOD_platelet, or_LOD_ANC, or_LOD_Neutropenia)
# Add p-value row
p_val_LOD_row <- c("p-value", p_val_LOD_Hb, p_val_LOD_WBC, p_val_LOD_Leukocytosis, p_val_LOD_Leukocytosis, p_val_LOD_platelet, p_val_LOD_ANC, p_val_LOD_Neutropenia)
# Bind both rows to the table
Table1_2 <- rbind(Table1_2, or_LOD_row, p_val_LOD_row)
# View the final table
Table1_2
```
## 1.3. ICU Admission  among infants
```{r}
# Logistic regression models
model_ICU_Hb            <- glm(ICU ~ Hb,data_clean_infants, family = binomial)
model_ICU_WBC           <- glm(ICU ~ WBC,data_clean_infants, family = binomial)
model_ICU_Leukocytosis  <- glm(ICU ~ Leukocytosis,data_clean_infants, family = binomial)
model_ICU_Leukopenia    <- glm(ICU ~ Leukopenia,data_clean_infants, family = binomial)
model_ICU_platelet      <- glm(ICU ~ platelet,data_clean_infants, family = binomial)
model_ICU_ANC           <- glm(ICU ~ ANC,data_clean_infants, family = binomial)
model_ICU_Neutropenia   <- glm(ICU ~ Neutropenia,data_clean_infants, family = binomial)
# Calculate OR and 95% CI for each factor
or_ICU_Hb           <- calculate_or_ci(model_ICU_Hb)
or_ICU_WBC          <- calculate_or_ci(model_ICU_WBC)
or_ICU_Leukocytosis <- calculate_or_ci(model_ICU_Leukocytosis)
or_ICU_Leukopenia   <- calculate_or_ci(model_ICU_Leukopenia)
or_ICU_platelet     <- calculate_or_ci(model_ICU_platelet)
or_ICU_ANC          <- calculate_or_ci(model_ICU_ANC)
or_ICU_Neutropenia  <- calculate_or_ci(model_ICU_Neutropenia)
# Create a p-value vector from the model_ICUs
p_val_ICU_Hb            <- round(coef(summary(model_ICU_Hb))[2,4], 3)
p_val_ICU_WBC           <- round(coef(summary(model_ICU_WBC))[2,4], 3)
p_val_ICU_Leukocytosis  <- round(coef(summary(model_ICU_Leukocytosis))[2,4], 3)
p_val_ICU_Leukopenia    <- round(coef(summary(model_ICU_Leukopenia))[2,4], 3)
p_val_ICU_platelet      <- round(coef(summary(model_ICU_platelet))[2,4], 3)
p_val_ICU_ANC           <- round(coef(summary(model_ICU_ANC))[2,4], 3)
p_val_ICU_Neutropenia   <- round(coef(summary(model_ICU_Neutropenia))[2,4], 3)
# Create the summary table
Table1_3 <- data_clean_infants |>
  mutate(Admission_loc = factor(Admission_loc, levels = c("other", "ICU"))) |>
  group_by(Admission_loc) |>
  summarise(N = n(),
            Hb_mean_SD = paste(round(mean(Hb, na.rm = TRUE), 2), " (", round(sd(Hb, na.rm = TRUE), 2), ")", sep = ""),
            WBC_mean_SD = paste(round(mean(WBC, na.rm = TRUE), 2), " (", round(sd(WBC, na.rm = TRUE), 2), ")", sep = ""),
            Leukocytosis_Yes_p = paste(sum(Leukocytosis == 1, na.rm = TRUE), " (", round(100 * mean(Leukocytosis == 1, na.rm = TRUE), 1), ")", sep = ""),
            Leukopenia_Yes_p = paste(sum(Leukopenia == 1, na.rm = TRUE), " (", round(100 * mean(Leukopenia == 1, na.rm = TRUE), 1), ")", sep = ""),
            Platelet_mean_SD = paste(round(mean(platelet, na.rm = TRUE), 2), " (", round(sd(platelet, na.rm = TRUE), 2), ")", sep = ""),
            ANC_mean_SD = paste(round(mean(ANC, na.rm = TRUE), 2), " (", round(sd(ANC, na.rm = TRUE), 2), ")", sep = ""),
            Neutropenia_Yes_p = paste(sum(Neutropenia == 1, na.rm = TRUE), " (", round(100 * mean(Neutropenia == 1, na.rm = TRUE), 1), ")", sep = "")) |>
  mutate(Value = paste0(Admission_loc, "(N=", N, ")"), .before = Hb_mean_SD) |>
  select(-Admission_loc, -N)
# Add OR row directly during table creation
or_ICU_row <- c("OR (95% CI)", or_ICU_Hb, or_ICU_WBC, or_ICU_Leukocytosis, or_ICU_Leukopenia, or_ICU_platelet, or_ICU_platelet, or_ICU_Neutropenia)
# Add p-value row
p_val_ICU_row <- c("p-value", p_val_ICU_Hb, p_val_ICU_WBC, p_val_ICU_Leukocytosis, p_val_ICU_Leukocytosis, p_val_ICU_platelet, p_val_ICU_ANC, p_val_ICU_Neutropenia)
# Bind both rows to the table
Table1_3 <- rbind(Table1_3, or_ICU_row, p_val_ICU_row)
# View the final table
Table1_3
```
## 1.4. Meningitis among infants
```{r}
# Logistic regression models
model_Meningitis_Hb            <- glm(Meningitis ~ Hb,data_clean_infants, family = binomial)
model_Meningitis_WBC           <- glm(Meningitis ~ WBC,data_clean_infants, family = binomial)
model_Meningitis_Leukocytosis  <- glm(Meningitis ~ Leukocytosis,data_clean_infants, family = binomial)
model_Meningitis_Leukopenia    <- glm(Meningitis ~ Leukopenia,data_clean_infants, family = binomial)
model_Meningitis_platelet      <- glm(Meningitis ~ platelet,data_clean_infants, family = binomial)
model_Meningitis_ANC           <- glm(Meningitis ~ ANC,data_clean_infants, family = binomial)
model_Meningitis_Neutropenia   <- glm(Meningitis ~ Neutropenia,data_clean_infants, family = binomial)
# Calculate OR and 95% CI for each factor
or_Meningitis_Hb           <- calculate_or_ci(model_Meningitis_Hb)
or_Meningitis_WBC          <- calculate_or_ci(model_Meningitis_WBC)
or_Meningitis_Leukocytosis <- calculate_or_ci(model_Meningitis_Leukocytosis)
or_Meningitis_Leukopenia   <- calculate_or_ci(model_Meningitis_Leukopenia)
or_Meningitis_platelet     <- calculate_or_ci(model_Meningitis_platelet)
or_Meningitis_ANC          <- calculate_or_ci(model_Meningitis_ANC)
or_Meningitis_Neutropenia  <- calculate_or_ci(model_Meningitis_Neutropenia)
# Create a p-value vector from the model_Meningitiss
p_val_Meningitis_Hb            <- round(coef(summary(model_Meningitis_Hb))[2,4], 3)
p_val_Meningitis_WBC           <- round(coef(summary(model_Meningitis_WBC))[2,4], 3)
p_val_Meningitis_Leukocytosis  <- round(coef(summary(model_Meningitis_Leukocytosis))[2,4], 3)
p_val_Meningitis_Leukopenia    <- round(coef(summary(model_Meningitis_Leukopenia))[2,4], 3)
p_val_Meningitis_platelet      <- round(coef(summary(model_Meningitis_platelet))[2,4], 3)
p_val_Meningitis_ANC           <- round(coef(summary(model_Meningitis_ANC))[2,4], 3)
p_val_Meningitis_Neutropenia   <- round(coef(summary(model_Meningitis_Neutropenia))[2,4], 3)
# Create the summary table
Table1_4 <- data_clean_infants |>
  mutate(Meningitis = factor(Meningitis, levels = c(0, 1))) |>
  group_by(Meningitis) |>
  summarise(N = n(),
            Hb_mean_SD = paste(round(mean(Hb, na.rm = TRUE), 2), " (", round(sd(Hb, na.rm = TRUE), 2), ")", sep = ""),
            WBC_mean_SD = paste(round(mean(WBC, na.rm = TRUE), 2), " (", round(sd(WBC, na.rm = TRUE), 2), ")", sep = ""),
            Leukocytosis_Yes_p = paste(sum(Leukocytosis == 1, na.rm = TRUE), " (", round(100 * mean(Leukocytosis == 1, na.rm = TRUE), 1), ")", sep = ""),
            Leukopenia_Yes_p = paste(sum(Leukopenia == 1, na.rm = TRUE), " (", round(100 * mean(Leukopenia == 1, na.rm = TRUE), 1), ")", sep = ""),
            Platelet_mean_SD = paste(round(mean(platelet, na.rm = TRUE), 2), " (", round(sd(platelet, na.rm = TRUE), 2), ")", sep = ""),
            ANC_mean_SD = paste(round(mean(ANC, na.rm = TRUE), 2), " (", round(sd(ANC, na.rm = TRUE), 2), ")", sep = ""),
            Neutropenia_Yes_p = paste(sum(Neutropenia == 1, na.rm = TRUE), " (", round(100 * mean(Neutropenia == 1, na.rm = TRUE), 1), ")", sep = "")) |>
  mutate(Value = paste0(Meningitis, "(N=", N, ")"), .before = Hb_mean_SD) |>
  select(-Meningitis, -N)
# Add OR row directly during table creation
or_Meningitis_row <- c("OR (95% CI)", or_Meningitis_Hb, or_Meningitis_WBC, or_Meningitis_Leukocytosis, or_Meningitis_Leukopenia, or_Meningitis_platelet, or_Meningitis_platelet, or_Meningitis_Neutropenia)
# Add p-value row
p_val_Meningitis_row <- c("p-value", p_val_Meningitis_Hb, p_val_Meningitis_WBC, p_val_Meningitis_Leukocytosis, p_val_Meningitis_Leukocytosis, p_val_Meningitis_platelet, p_val_Meningitis_ANC, p_val_Meningitis_Neutropenia)
# Bind both rows to the table
Table1_4 <- rbind(Table1_4, or_Meningitis_row, p_val_Meningitis_row)
# View the final table
Table1_4
```

## Table 1
```{r}
# infants vs older individuals

## P-values
p_val_Hb            <- round(coef(summary(glm(infant ~ Hb, data_clean, family = binomial)))[2,4], 3)
p_val_WBC           <- round(coef(summary(glm(infant ~ WBC, data_clean, family = binomial)))[2,4], 3)
p_val_Leukocytosis  <- round(coef(summary(glm(infant ~ Leukocytosis, data_clean, family = binomial)))[2,4], 3)
p_val_Leukopenia    <- round(coef(summary(glm(infant ~ Leukopenia, data_clean, family = binomial)))[2,4], 3)
p_val_platelet      <- round(coef(summary(glm(infant ~ platelet, data_clean, family = binomial)))[2,4], 3)
p_val_ANC           <- round(coef(summary(glm(infant ~ ANC, data_clean, family = binomial)))[2,4], 3)
p_val_Neutropenia   <- round(coef(summary(glm(infant ~ Neutropenia, data_clean, family = binomial)))[2,4], 3)
## Table
Table1_1 <- data_clean |> 
  group_by(infant)|>
  summarise(N=n(), 
            Hb_mean_SD= paste(round(mean(Hb, na.rm=T),2), " (", round(sd(Hb, na.rm=T), 2),")", sep=""), 
            WBC_mean_SD= paste(round(mean(WBC, na.rm=T),2), " (", round(sd(WBC, na.rm=T),2),")", sep=""), 
            Leukocytosis_Yes_p = paste(sum(Leukocytosis==1, na.rm = T), " (", round(100*mean(Leukocytosis==1, na.rm = T), 1),")", sep=""), 
            Leukopenia_Yes_p  = paste(sum(Leukopenia==1, na.rm = T), " (", round(100*mean(Leukopenia==1, na.rm = T), 1),")", sep=""), 
            Platelet_mean_SD= paste(round(mean(platelet, na.rm=T),2), " (",  round(sd(platelet, na.rm=T),2),")", sep=""), 
            ANC_mean_SD= paste(round(mean(ANC, na.rm=T),2), " (",  round(sd(ANC, na.rm=T),2),")", sep=""), 
            Neutropenia_Yes_p = paste(sum(Neutropenia==1, na.rm = T), " (",  round(100*mean(Neutropenia==1, na.rm = T),1),")", sep="")) |>
  mutate(Value= paste0(infant, "(N=", N, ")"), .before =Hb_mean_SD)|>
  select(-infant, -N)
Table1_1 <- rbind(Table1_1, c("p-value", p_val_Hb, p_val_WBC, p_val_Leukocytosis, p_val_Leukopenia, p_val_platelet, p_val_ANC,p_val_Neutropenia))



# ICU vs other  
## P-values
p_val_3_Hb <- round(wilcox.test(Hb ~ Meningitis, data = data_clean|> filter(days_at_dx <= 365))$p.value, 3)
p_val_3_WBC <- round(wilcox.test(WBC ~ Meningitis, data = data_clean|> filter(days_at_dx <= 365))$p.value, 3)
p_val_3_Leukocytosis <- round(coef(summary(glm(Leukocytosis ~ Meningitis, data_clean|> filter(days_at_dx <= 365), family = binomial)))[2,4], 3)
p_val_3_Leukopenia <- round(coef(summary(glm(Leukopenia ~ Meningitis, data_clean|> filter(days_at_dx <= 365), family = binomial)))[2,4], 3)
p_val_3_platelet <- round(wilcox.test(platelet ~ Meningitis, data = data_clean|> filter(days_at_dx <= 365))$p.value, 3)
p_val_3_ANC <- round(wilcox.test(ANC ~ Meningitis, data = data_clean|> filter(days_at_dx <= 365))$p.value, 3)
p_val_3_Neutropenia <- round(coef(summary(glm(Neutropenia ~ Meningitis, data_clean|> filter(days_at_dx <= 365), family = binomial)))[2,4], 3)
## Table
Table1_3 <- data_clean |> 
  filter(days_at_dx <= 365)|>
  group_by(Meningitis)|>
  summarise(N=n(), 
            Hb_mean_SD= paste(round(mean(Hb, na.rm=T),2), " (", round(sd(Hb, na.rm=T), 2),")", sep=""), 
            WBC_mean_SD= paste(round(mean(WBC, na.rm=T),2), " (", round(sd(WBC, na.rm=T),2),")", sep=""), 
            Leukocytosis_Yes_p = paste(sum(Leukocytosis==1, na.rm = T), " (", round(100*mean(Leukocytosis==1, na.rm = T), 1),")", sep=""), 
            Leukopenia_Yes_p  = paste(sum(Leukopenia==1, na.rm = T), " (", round(100*mean(Leukopenia==1, na.rm = T), 1),")", sep=""), 
            Platelet_mean_SD= paste(round(mean(platelet, na.rm=T),2), " (",  round(sd(platelet, na.rm=T),2),")", sep=""), 
            ANC_mean_SD= paste(round(mean(ANC, na.rm=T),2), " (",  round(sd(ANC, na.rm=T),2),")", sep=""), 
            Neutropenia_Yes_p = paste(sum(Neutropenia==1, na.rm = T), " (",  round(100*mean(Neutropenia==1, na.rm = T),1),")", sep="")) |>
  mutate(Value= paste0(Meningitis, "(N=", N, ")"), .before =Hb_mean_SD)|>
  select(-Meningitis, -N)
Table1_3 <- rbind(Table1_3, c("p-value", p_val_3_Hb, p_val_3_WBC, p_val_3_Leukocytosis, p_val_3_Leukopenia, p_val_3_platelet, p_val_3_ANC,p_val_3_Neutropenia))
  
rbind(Table1_1, Table1_2, Table1_3)
```

# Identifying correlates of severe clinical outcomes
# Function
```{r}
# Generalized function to calculate OR and adjusted OR controlling for CC
calculate_or <- function(df, risk_factor, outcome) {
  
  # Ensure that risk_factor is treated as a column from the dataframe
  risk_factor_col <- df[[risk_factor]]
  
  # Create a 2x2 table of the outcome vs. Other by the risk factor
  outcome_cases <- sum(df |> filter(df[[outcome]] == 1) |> pull(risk_factor))
  other_cases <- sum(df |> filter(df[[outcome]] == 0) |> pull(risk_factor)) 
  
  # Logistic regression without controlling for CC (unadjusted OR)
  log_reg <- glm(as.formula(paste(outcome, "~", risk_factor)), data = df, family = binomial)
  log_reg_summary <- coef(summary(log_reg))
  
  # Extract Odds Ratio (exponentiate the coefficient) and Confidence Intervals for unadjusted OR
  or_value <- exp(log_reg_summary[2, 1])
  ci_lower <- exp(log_reg_summary[2, 1] - 1.96 * log_reg_summary[2, 2])
  ci_upper <- exp(log_reg_summary[2, 1] + 1.96 * log_reg_summary[2, 2])
  p_value <- round(log_reg_summary[2, 4], 3)
  
  # Logistic regression controlling for CC (adjusted OR)
  log_reg_adj <- glm(as.formula(paste(outcome, "~", risk_factor, "+ CC")), data = df, family = binomial)
  log_reg_adj_summary <- coef(summary(log_reg_adj))
  
  # Extract Odds Ratio (exponentiate the coefficient) and Confidence Intervals for adjusted OR
  adj_or_value <- exp(log_reg_adj_summary[2, 1])
  adj_ci_lower <- exp(log_reg_adj_summary[2, 1] - 1.96 * log_reg_adj_summary[2, 2])
  adj_ci_upper <- exp(log_reg_adj_summary[2, 1] + 1.96 * log_reg_adj_summary[2, 2])
  adj_p_value <- round(log_reg_adj_summary[2, 4], 3)
  
  # Return a dataframe with both OR and adjusted OR
  data.frame(
    Risk_Factor = risk_factor,
    outcome_yes = outcome_cases,
    outcome_No = other_cases,
    Crude_Or = sprintf("%.2f (%.2f - %.2f)", or_value, ci_lower, ci_upper),
    P_value = p_value,
    Adjusted_Or = sprintf("%.2f (%.2f - %.2f)", adj_or_value, adj_ci_lower, adj_ci_upper),
    Adjusted_Pvalue = adj_p_value
  )
}

# List of risk factors (genes) to test
risk_factors <- c("Surface_protein.ALP1", 
                  "Surface_protein.ALP23",
                  "Surface_protein.ALPHA",
                  "Surface_protein.RIB",
                  "Surface_protein.PI1" ,
                  "Surface_protein.PI2A1" ,
                  "Surface_protein.PI2A2",
                    "Surface_protein.PI2B",
                      "Surface_protein.SRR1",
                     "Surface_protein.SRR2",
                     "Surface_protein.Sip.1a",
                  "Surface_protein.Sip.3a",
                  "Surface_protein.HVGA",
                  "Virulence_Gene.lmb",
                  "Virulence_Gene.scpB",
                  "Virulence_Gene.hylB",
                  "Virulence_Gene.fbsB" )


```

# TABLE 2   
# Outcome : ICU
```{r}
# Calculate OR, CI, and P-value for each gene, and adjust for CC
results <- lapply(risk_factors, calculate_or, df = data_clean_infants, outcome = "ICU")

# Combine the results into a single dataframe
final_table_ICU <- bind_rows(results) |> 
  mutate(Risk_Factor = str_remove(str_remove(Risk_Factor, "Surface_protein."), "Virulence_Gene."))|>
  mutate(sig = ifelse(Adjusted_Pvalue < round(0.05/length(risk_factors), 3), "yes", "no"))

# Rename columns for clarity
colnames(final_table_ICU) <- c("Risk_Factor", "ICU (n=31)","Other (n=39)","Crude OR (95% CI)", "P-value", "Adjusted OR (95% CI)","Adjusted P-value", "Significant")

final_table_ICU
```
# Outcome : Meningitis
```{r}
# Calculate OR, CI, and P-value for each gene, and adjust for CC
results <- lapply(risk_factors, calculate_or, df = data_clean_infants, outcome = "Meningitis")

# Combine the results into a single dataframe
final_table_Meningitis <- bind_rows(results) |> 
  mutate(Risk_Factor = str_remove(str_remove(Risk_Factor, "Surface_protein."), "Virulence_Gene."))|>
  mutate(sig = ifelse(Adjusted_Pvalue< round(0.05/length(risk_factors), 3), "yes", "no"))

# Rename columns for clarity
colnames(final_table_Meningitis) <- c("Risk_Factor", "Meningitis (n=13)","No Meningitis (n=57)","Crude OR (95% CI)", "P-value", "Adjusted OR (95% CI)","Adjusted P-value", "Significant")

final_table_Meningitis
```

# Outcome : Age of onset 
## LOD vs VLOD among infants
```{r}
# Subset data
data_LOD <- data_clean_infants |> filter(Age_cat != "EOD") |> mutate(LOD= ifelse(Age_cat == "LOD", 1, 0))

# Calculate OR, CI, and P-value for each gene, and adjust for CC
results <- lapply(risk_factors, calculate_or, df = data_LOD, outcome = "LOD")

# Combine the results into a single dataframe
final_table_LOD <- bind_rows(results) |> 
  mutate(Risk_Factor = str_remove(str_remove(Risk_Factor, "Surface_protein."), "Virulence_Gene."))|>
  mutate(sig = ifelse(Adjusted_Pvalue< round(0.05/length(risk_factors), 3), "yes", "no"))

# Rename columns for clarity
colnames(final_table_LOD) <- c("Risk_Factor", "LOD (n=48)","VLOD (n=20)","Crude OR (95% CI)", "P-value", "Adjusted OR (95% CI)","Adjusted P-value", "Significant")

final_table_LOD
```
## infant vs older individuals
```{r}
# Calculate OR, CI, and P-value for each gene, and adjust for CC
results <- lapply(risk_factors, calculate_or, df = data_clean, outcome = "infant")

# Combine the results into a single dataframe
final_table_infant <- bind_rows(results) |> 
  mutate(Risk_Factor = str_remove(str_remove(Risk_Factor, "Surface_protein."), "Virulence_Gene."))|>
  mutate(sig = ifelse(Adjusted_Pvalue< round(0.05/length(risk_factors), 3), "yes", "no"))


# Rename columns for clarity
colnames(final_table_infant) <- c("Risk_Factor", "Infant (n=70)","Older (n=17)","Crude OR (95% CI)", "P-value", "Adjusted OR (95% CI)","Adjusted P-value", "Significant")

final_table_infant
```

## Neutropenia
```{r}
# Calculate OR, CI, and P-value for each gene, and adjust for CC
results <- lapply(risk_factors, calculate_or, df = data_clean_infants, outcome = "Neutropenia")

# Combine the results into a single dataframe
final_table_Neutropenia <- bind_rows(results) |> 
  mutate(Risk_Factor = str_remove(str_remove(Risk_Factor, "Surface_protein."), "Virulence_Gene."))|>
  mutate(sig = ifelse(Adjusted_Pvalue< round(0.05/length(risk_factors), 3), "yes", "no"))

# Rename columns for clarity
colnames(final_table_Neutropenia) <- c("Risk_Factor", "Neutropenia (n=17)","No Neutropenia (n=51)","Crude OR (95% CI)", "P-value", "Adjusted OR (95% CI)","Adjusted P-value", "Significant")

final_table_Neutropenia
```

## Leukopenia
```{r}
# Calculate OR, CI, and P-value for each gene, and adjust for CC
results <- lapply(risk_factors, calculate_or, df = data_clean_infants, outcome = "Leukopenia")

# Combine the results into a single dataframe
final_table_Leukopenia <- bind_rows(results) |> 
  mutate(Risk_Factor = str_remove(str_remove(Risk_Factor, "Surface_protein."), "Virulence_Gene."))|>
  mutate(sig = ifelse(Adjusted_Pvalue< round(0.05/length(risk_factors), 3), "yes", "no"))

# Rename columns for clarity
colnames(final_table_Leukopenia) <- c("Risk_Factor", "Leukopenia (n=24)","No Leukopenia (n=44)","Crude OR (95% CI)", "P-value", "Adjusted OR (95% CI)","Adjusted P-value", "Significant")

final_table_Leukopenia

data_clean_infants |> group_by(Leukopenia) |> summarise(N=n())

```

## Leukocytosis
```{r}
# Calculate OR, CI, and P-value for each gene, and adjust for CC
results <- lapply(risk_factors, calculate_or, df = data_clean_infants, outcome = "Leukocytosis")

# Combine the results into a single dataframe
final_table_Leukocytosis <- bind_rows(results) |> 
  mutate(Risk_Factor = str_remove(str_remove(Risk_Factor, "Surface_protein."), "Virulence_Gene."))|>
  mutate(sig = ifelse(Adjusted_Pvalue< round(0.05/length(risk_factors), 3), "yes", "no"))

# Rename columns for clarity
colnames(final_table_Leukocytosis) <- c("Risk_Factor", "Leukocytosis (n=24)","No Leukocytosis (n=44)","Crude OR (95% CI)", "P-value", "Adjusted OR (95% CI)","Adjusted P-value", "Significant")

final_table_Leukocytosis

data_clean_infants |> group_by(Leukocytosis) |> summarise(N=n())
data_clean_infants |> mutate(NEW_Leukocytosis = ifelse(WBC > 10, 1, 0)) |> group_by(NEW_Leukocytosis) |> summarise(N=n())
```
